plyranges: a grammar for transforming genomics data

### 2018-03-21 paper.Rmd
more changes to intro

In the Bioconductor/R packages `IRanges` and `GenomicRanges` these 
representations have been implemented as a suite of data structures called
_Ranges_ @Lawrence2013-wg; @Huber2015-ei. These data structures cover 
many common data types  encountered in bioinformatics analyses.

The Bioconductor infrastructure for computing with genomic ranges are highly 
efficient and powerful, however the application programming interface (API) for 
performing analysis tasks with _Ranges_ is complex due to its large number of
methods and classes. It also makes resulting scripts written 
difficult for a non-programmer to read and reason about. An alternative approach 
would be to implement a domain specific language (DSL) as a fluent interface 
built on top _Ranges_ [@Fowler2010-zd]. The goal of a fluent interface is to 
enable users to write  human-readable code via method chaining and consistent function returns. Fluent  interfaces fit naturally in the context of bioinformatics workflows because they 
enable writing succinct pipelines.

Several other DSLs have been implemented to reason about genomics data. Broadly, 
these are either implemented as query languages or as command line tools 
embedded in the Unix environment.  An example of the former is the 
Genome Query Language (GQL) and its distributed 
implementation GenAp which use an SQL-like syntax for fast retrieval of 
information from genomic databases and BAM files @Kozanitis2014-va; @Kozanitis2016-bm. 
Another  example is the Genometric Query Language (GMQL) which implements a 
relational algebra for combining big genomic datasets @Kaitoua2017-pw.  
The command line application 
BEDtools develops an extensive algebra for performing arithmetic between two or 
more sets of genomic regions  @Quinlan2010-gc. It also has a python interface 
which simplifies constructing scripts for performing analyses based on BEDTools 
@Dale2011-js. 
<!--- probably should also mention something about their cons, and more detail --->

The abstraction provided by the _Ranges_ data structures aligns with the concept
of tidy data @Wickham2014-jc. A tidy dataset is defined to be to a tabular
data structure that has observations as rows and columns as variables, and each
tidy dataset represents measurements from a single observational unit.
The tidy data pattern is useful because it allows 
us to see how the data relates to the design of an experiment
and the variables measured. Consequently, it makes both the modelling and 
manipulation of data more systematic. The _Ranges_ data structure follows 
this abstraction: it is a rectangular table corresponding to a single biological 
context. Each row contains a single observation and each column is a variable about
that observation.

The _plyranges_ API implements a domain specific language using the existing
_IRanges_ and _GenomicRanges_ packages in Bioconductor as a back-end. Consequently,
our API still has the speed and efficiency of the aforementioned packages but
with a more coherent interface. The API also
extends the grammar elements in _dplyr_ [@Wickham2017-dplyr] 
for performing genomic specific
manipulations such as finding overlapping regions or nearest neighbour regions
between many _Ranges_ . The _plyranges_ API is specifically designed to enable
fast interactive analysis of _Ranges_ objects but can also be used for scripting 
genomic data workflows.

>>>>

Many domain specific languages (DSLs) have been created to process and reason
about heterogenous genomics data. Broadly, they are implemented as either
query languages or as an embedded command line tool.
An example of the former is the Genome Query Language (GQL) and its distributed 
implementation GenAp which use an SQL-like syntax for fast retrieval of 
information of unprocessed sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. 
Similarly, the Genometric Query Language (GMQL) implements a 
relational algebra for combining genomic datasets @Kaitoua2017-pw.  
The command line application BEDtools develops an extensive algebra for 
performing arithmetic between two or more sets of genomic regions 
@Quinlan2010-gc; @Dale2011-js. All of the aformentioned DSLs are designed to 
be run either at the command line or embedded in scripts for batch processing. 
As a consquence, they are not integrated into an environment designed to 
perform interactive data  analysis that has access to state of the art 
methods for statistics and visualisation like the R language.

The Bioconductor/R packages `IRanges` and `GenomicRanges` 
define a DSL for analysing genomics data with R via
an object-oriented (OO) interface @Lawrence2013-wg; @Huber2015-ei. This interface
provides abstractions for representing genomics data and enables interoperability
by allowing users and developers to use these abstractions in their own code
and packages. However, due to the large number of abstractions there are costs
involved. Due to the OO interface each abstraction will behave slightly 
differently for any given generic method. There is a cognitive load on the user
to understand the appropriate context for each abstraction. An alternative approach
is to recognise that most genomic data analysis tasks can be performed with 
data structures that naturally represent genomic intervals.

This natural abstraction aligns with the concept of tidy data @Wickham2014-jc. 
A tidy dataset is defined to be to a tabular
data structure that has observations as rows and columns as variables, and each
tidy dataset represents measurements from a single observational unit.
The tidy data pattern is useful because it allows 
us to see how the data relates to the design of an experiment
and the variables measured. The genomic interval data structure follows 
this pattern: it is a rectangular table corresponding to a single biological 
context. Each row contains a single observation and each column is a variable about
that observation.

The tidy data abstraction enables the construction of fluent application programming interfaces (APIs) [@Fowler2010-zd] that are optimised for readability. An example of a 
fluent interface built on the tidy data pattern is the R pacakge `dplyr` [@Wickham2017-dplyr]. 
The package defines a consistent syntax and style for transforming tabular data sets via
functions that are named as verbs. Each function performs an action on the data
set and is cohesive and endomorphic. Cohesion means that each function performs
a singular task and has no side-effects, while endomorphism means a function
call returns the same type of object as its input.

Then the last paragraph (already pretty close) states that:

    we aim to simplify the interface to the most fundamental Ranges data structures, which we must reuse for the reasons stated above,
    it is possible to align the Ranges data structures with the tabular notions of the dplyr API
    so we will extend dplyr to incorporate genomic notions

We have created a DSL called `plyranges` that uses the existing
`IRanges` and `GenomicRanges` packages in Bioconductor as a back-end. Our API 
still has the speed and efficiency of these packages but
with a consistent interface. The API also extends the grammar elements in `dplyr` 
for performing genomic specific manipulations such as finding overlapping 
regions or nearest neighbour regions between genomic intervals. 
The _plyranges_ DSL is specifically designed to enable fast interactive 
analysis of _Ranges_ objects but can also be used for scripting 
genomic data workflows.
